Future challenges for Research Software Engineering

### 2025-05-19 future_challenges_for_RSEng.tex
update story

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
climate \cite{Lannelongue2023} and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large, foreign corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay used to work on a project that developed
software for cheap water monitoring devices in the Global South, but now funding has dried up.
Having gone through another fixed term contract he is now fed up
with work in academia and decides that its time limited contracts are not attractive.
Now Governments around the world are struggling to reposition themselves in this emerging de-globalised world with its new blocks and therefore have de-prioritised publicly funded research.
Kay sees the writings on the wall and decides to work for a defence company since they offer more secure employment in this polarised world.
\end{story}

>>>>

\subsection{Impact of AI and automation}
\label{sec:ai}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering \cite{s2erc,siadati2024},
climate \cite{Lannelongue2023} and more.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.
With the advantages of \autoref{sec:ai} in place, ever-more AI created artifacts will be in the world.
Therefore we expect a discussion within societies about their acceptable use to take place.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large, foreign corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment. There he gained valuable experience working in an industrial setting. However, he was very happy when an opportunity arose to work in a central RSE department with an open-ended contract. Apart from his main responsibility of developing and maintaining software for an affordable water monitoring system used in the global south, he now also teaches undergraduates digital literacy.
\end{story}
